An introduction to Docker for reproducible research

### 2014-11-05 docker-paper/manuscript.Rmd
texlive example

I introduce the most relevant concepts from Docker through the 
context of the four challenges for reproducible research I have
discussed above.

@@

A key difference between Docker images and other virtual machines is 
that the Docker images share the Linux kernel with the host machine.
For the end user the primary consequence of this is that any Docker
image must be based on a Linux system with Linux-compatible software, which
includes (R, Python, Matlab, and most other scientific programming needs). [^4]

Sharing the Linux kernel makes Docker much more light-weight and higher
performing than complete virtual machines -- a typical desktop computer
could run no more than a few virtual machines at once but 
would have no trouble running 100's of Docker containers (a container
is simply the term for running instance of an image). This feature
has made Docker particularly attractive to industry and is largely
responsible for the immense popularity of Docker. For our purposes this
is a nice bonus, but the chief value to reproducible research lies
in other aspects.

@@

To address this concern, one must archive a binary copy of the image
used at the time the research was first performed.  Docker provides a
simple utility to save an image as a portable `tarball` file that can
be read in by any other Docker installation, providing a robust way to
run the exact versions of all software involved.  By testing both the
`tarball` archive and the image generated by the latest Dockerfile,
Docker provides a simple way to confirm whether or not code rot has
effected the function of a particular piece of code.

@@

Though Docker is not immune to these challenges, it offers an interesting 
example of a way forward in addressing these fundamental concerns. Here 
I highlight five of these features in turn:

- Local environment
- Modular reuse
- Portable environments 
- Public repository for sharing
- Versioning

## Using Docker as a local development environment

Perhaps the most important feature of a reproducible research tool is that
it be easy to learn and fit relatively seamlessly into existing workflow
patterns of domain researchers.  This, more than any other concern,
can explain the relatively low uptake of previously proposed solutions.
Being a new and unfamiliar tool to most domain scientists, Docker is
far from immune to the same critique.  Nevertheless, Docker takes us
several key steps towards an approach that can be easily adopted in a
research context.

<!-- Part of the reason is better image distribution-->
While proponents of virtual machines for reproducible research propose
that these machines would be available exclusively as cloud computing
environments [_e.g._ @Dudley_2010], many researchers work _locally_, that
is, primarily with software that is installed on their laptop or desktop
computer, and turning to cloud-based or other remote platforms only for
certain collaborative tasks or when the work is mature enough to need
increased computational power.

rely more on the graphic interface tools for tasks such as managing
files, text editing, debugging, IDEs, or interacting with version
control systems.  Scientific computing on remote machines, by contrast,
still relies largely on potentially less familiar text based command
line functions for these tasks (though web based interfaces like RStudio
Server are rapidly filling this gap).

Docker can be easily installed on most major platforms (see
[https://docs.Docker.com/installation](https://docs.docker.com/installation);
On systems not already based on the Linux Kernel, such as Mac or Windows,
this is accomplished through the use of a small VirtualBox-based VM
running on the host OS called `boot2docker`) and run locally.  Docker
allows a user to link any directory (such as the working directory
for a particular script or project) to the running Docker container.
This allows a user to rely on the familiar tools of the host OS for
roles such as text editing, file browsing, or version control, while
still allowing code execution to occur inside the controlled development
environment of the container.

For example, one launches an interactive R console in a container that
is linked to our current working directory like so:

```bash
docker run -v $(pwd):/ -it cboettig/rstudio /usr/bin/R
```

The resulting system behaves almost identically[^2] to running R on
the command line of the host OS. (Clearly a similar command could be
used just as well with interactive shells from other languages such as
`ipython` or `irb`).

[^2]: External windows, such graphics windows cannot be opened directly
from the container in this setup.  Graphics would have to be saved to
disk as raster or vector files and viewed on the host OS, until a better
solution can be found. RStudio Server ([rstudio.com/products/rstudio/#Server](http://rstudio.com/products/rstudio/#Server)) is one way around this.

<!--

### Using RStudio Server

An alternative approach for working locally with familiar tools is to
leverage web-based clients such as RStudio. 
In the R environment, the open source RStudio IDE provides
another key component in making this system accessible to most
domain scientists. Because the RStudio IDE is written completely with
standard web languages (Javascript, CSS, and HTML), its web-based 
RStudio Server looks and feels identical to its popular desktop IDE.
RStudio Server provides a way to interact with R on a remote environment
without the latency, X tunnelling, and so forth typically involved in running R
on a remote server.

RStudio server provides users a way to interact with R,
the file system, git, text editors, and graphics running on a Docker
container.  Users already familiar with the popular IDE can thus benefit
from the reproducibility and portability features provided by running R
in a container environment without having to adapt to a new workflow.

To accompany this paper, I provide a Docker image for running RStudio
server, which can be launched as follows.  From the terminal (or
boot2docker terminal on Mac or Windows client), run:

```bash
sudo docker run -d -p 8787:8787 cboettig/ropensci
```

That will take a while to download the image the first time you run it. A
boot2docker user will then need to determine the ip address assigned
to boot2docker as shown below, while a Linux user can just specify
`http://localhost`.

```bash
boot2docker ip
```

Add the port `:8787` to the end of this address and paste
it into your browser address bar, which should open to the
RStudio welcome screen.  A user can now login with user/password
`rstudio/rstudio`, run R scripts, install packages, use git,
and so forth. User login and other configurations can be
customized using environmental variables; see details at
[https://github.com/ropensci/Docker](https://github.com/ropensci/docker)

@@

The approach of Docker offers a technical
solution to what is frequently seen as the primary weakness of the
standard virtual machine approach to reproducibility - reusing and
remixing elements.  To some extent this is already addressed by the
DevOps approach of Dockerfiles, providing a scripted description
of the environment that can be tweaked and altered, but also includes
something much more fundamental to Docker.

The challenge to reusing virtual machines can be summarized as "you
can't install an image for every pipeline you want..." [@Watson_2012]. In
contrast, this is exactly how Docker containers are designed to work.
There are at least two ways in which Docker supports this kind of
extensibility.

First, Docker facilitates modular reuse by building one container on top of
another.  Rather than copy a Dockerfile and then start adding more lines
to the bottom, one can declare a new Dockerfile is built on an old one
using the `FROM` directive. Here is an example Dockerfile that adds the
R statistical computing environment to a basic Ubuntu Linux distribution.

```
FROM ubuntu:14.04
RUN apt-get update 
RUN apt-get -y install r-recommended
```

This acts like a software dependency; but unlike other software, a
Dockerfile must have exactly one dependency (one `FROM` line). Note
that a particular version of the dependency can be specified using the
`:` notation (similarly, package managers like `apt` show here can be
directed to install specific versions, if desired. One can in fact be much
more precise by specific cryptographic hash of a particular binary image.
In practice, using hashes in FROM directives may be undesirable as it also
avoids adopting security updates that may have been made to the base image.

One can name and build this Dockerfile by running in the
working directory:

```bash
docker build -t username/r-recommended .
```

The key point here is that other researchers can easily build off this
image just created, extending our work directly, rather than having
to go back to the original image.

If this image is shared (either directly or through a Docker Hub),
another user can now build directly on our image, rather than on the
`ubuntu:14.04` image used as a base in the first Dockerfile.  To do
so, one simply specifies a different image in the `FROM` directive of
the Dockerfile, followed by the lines required to add any additional
software required.

```
FROM username/r-recommended 
RUN apt-get update
RUN apt-get -y install r-cran-matrix
```

Though this shows only very minimal examples that add a single piece of
software in each step, clearly this approach can be particularly powerful
in building up more complex environments.

Each one acts as a building block providing just what is necessary to
run one particular service or element, and exposing just what is necessary
to link it together with other blocks.  For instance, one could have one
container running a PostgreSQL database which serves data to another container
running a python environment to analyze the data:

@@

Unlike the much more heavyweight virtual machine approach, containers
are implemented in way such that a single computer can easily run 100s
of such services each in their own container, making it easy to break
computational elements down into logically reusable chunks that come,
batteries included, with everything they need to run reproducibly.

>>>>

I introduce the most relevant concepts from Docker through the context
of the four challenges for reproducible research I have discussed above.
In brief, these challenges can be addressed by distributing a Dockerfile
capable of reconstructing the researcher's development environment,
ideally along with depositing the corresponding binary Docker image in
an appropriate repository.

@@

A key difference between Docker images and other virtual machines is 
that the Docker images share the Linux kernel with the host machine.
For the end user the primary consequence of this is that any Docker
image must be based on a Linux system with Linux-compatible software, which
includes R, Python, Matlab, and most other scientific programming needs.[^4]

Sharing the Linux kernel makes Docker more light-weight and higher
performing than complete virtual machines -- a typical desktop computer
could run no more than a few virtual machines at once but 
would have no trouble running 100's of Docker containers (a container
is simply the term for running instance of an image). This feature
has made Docker particularly attractive to industry and is largely
responsible for the immense popularity of Docker. For our purposes this
is a nice bonus, but the chief value to reproducible research lies
in other aspects.

@@

To address this concern, one must archive a binary copy of the image
used at the time the research was first performed.  Docker provides a
simple utility to save an image as a portable `tarball` file that can
be read in by any other Docker installation, providing a robust way to
run the exact versions of all software involved.  By testing both the
`tarball` archive and the image generated by the latest Dockerfile,
Docker provides a simple way to confirm whether or not code rot has
effected the function of a particular piece of code. Binary Docker images
can be efficiently shared through the Docker Hub as well, as I describe
later.

@@

Though Docker is not immune to these challenges, it offers an interesting 
example of a way forward in addressing these fundamental concerns. Here 
I highlight these features in turn:

- Integrating into local development environments
- Modular reuse
- Portable environments 
- Public repository for sharing
- Versioning

## Integrating into local development environments

Due in part to the difficulty of moving large VM images around a network,
proponents of virtual machines for reproducible research often propose
that these machines would be available exclusively as cloud computing
environments [_e.g._ @Dudley_2010] rather than downloaded to a user's
personal computer. Though cloud computing offers some advantages such as
scalable resources for computationally intensive tasks, many researchers
prefer to work with tools installed locally on their personal computer,
at least during testing and development.  This can reduce costs, issues of
network latency, the ability to work off-line, and will be most familiar
to students and new developers.

It is possible to run virtual machines locally on most common laptop and
desktop computers, as demonstrated in the approach now being pioneered
at UC Berkeley [@Clark_2014].  This approach provides users with a pixel-identical
environment whether working locally or on the cloud, which has particular
advantages for student instruction [@Clark_2014]. However, it remains to be
seen if existing researchers will be willing to forgo native applications
for such tasks as file browsing, text editing, or version management and 
rely exclusively on this standardized virtual environment.

In contrast, Docker's approach is optimized for a more integrated
workflow.  Rather than replace a user's existing toolchain with a
standardized virtual environment, editors and all, Docker is optimized at
the level of single applications. A developer can thus rely on familiar
tools while still ensuring that the execution of their code always occurs
on the standardized container environment, thus ensuring its portability
and reproducibility. This approach can be accomplished either by linking
volumes or directories between the container and the host, or simply
by instructing the Dockerfile to copy the code to the container (As we will
see, modular reuse makes this latter strategy efficient).

Rather than put the burden on the researcher to adopt
a very different workflow, the researcher can thus use their familiar editors,
etc, while immediately benefiting from the fact their computations can
be easily deployed on the cloud or the machines of collaborators without
any further effort than installing Docker software. The docker approach
is particularly well suited for moving between local and cloud platforms
when a web-based integrated development environment is available, such as
RStudio Server, or (to lesser extent) an iPython notebook.

On systems not already based on the Linux kernel (such
as Mac or Windows platforms), Docker is installed (see see
[docs.docker.com/installation](https://docs.docker.com/installation))
through means of a very small (about
24 MB) VM platform called `boot2docker`
([github.com/docker/boot2docker](https://github.com/docker/boot2docker)),.
While this poses an additional challenge for tight integration with desktop
tools, recent advances in Docker are rapidly bridging this gap as well. 
For example, Docker 1.3, released between drafts of this manuscript, supports
shared volumes on Macs ([blog.docker.com](http://blog.docker.com/2014/10/docker-1-3-signed-images-process-injection-security-options-mac-shared-directories/)).

@@

The approach of Docker offers a technical
solution to what is frequently seen as the primary weakness of the
standard VM approach to reproducibility - reusing and
remixing elements.  To some extent this is already addressed by the
DevOps approach of Dockerfiles, providing a scripted description
of the environment that can be tweaked and altered, but also includes
something much more fundamental to Docker.

The challenge to reusing VMs can be summarized as "you
can't install an image for every pipeline you want..." [@Watson_2012]. 
While providing a VM may make it easy for other researchers to run
a particular piece of software, it becomes very difficult to combine
multiple software components in future research if each must run inside
it's own VM. The lack of reusable, scalable modules in the VM model
poses a major barrier to future reuse.  In contrast, (as the analogy to 
shipping containers in the name might imply) Docker containers
are optimized for this kind of stacking and modular reuse.
There are at least three ways in which Docker supports this kind of
extensibility.

Most primitively, because Dockerfile itself provides a script for
creating the computational environment, future researchers can
extend or modify the resulting machine image by simply editing the
Dockerfile. This same approach is also possible for VM approaches that rely on 
DevOps tools [@Clark_2014]. In contrast to VMs, however, Docker is also
modular by design: both in how Dockerfiles are defined and how Docker
containers can be linked.

First, Docker facilitates modular reuse by building one container on top
of another through the use of `FROM` directive in Dockerfiles. This acts
like a software dependency; but unlike other software, a Dockerfile must
have exactly one dependency (one `FROM` line). Particular version of
the dependency can be specified using the `:` notation, or ommitted to
default to the latest version. To install software, Dockerfiles leverage
existing package managers on common Linux distributions such as `apt`
on Debian. These also permit installing either specific or only the
latest version.

Consider the following Dockerfile, which

```
FROM rocker/ropensci:latest
RUN git clone https://github.com/ropensci/RNeXML.git /rnexml
RUN Rscript -e 'devtools::install("/rnexml"); devtools::check("/rnexml")'
```

A Dockerfile can have only a single `FROM` line, while sometimes it may
be necessary to build on the computational environment provided by more
than one container. To address this, Docker defines a syntax for linking
multiple containers together:

Each one acts as a building block providing just what is necessary to
run one particular service or element, and exposing just what is necessary
to link it together with other blocks.

```bash
docker run --name tex -v /usr/local/texlive rocker/texlive
docker run -dP --volumes-from tex \
-e PATH=$PATH:/usr/local/texlive/2014/bin/x86_64-linux/ \
rocker/rstudio 
```

For instance, one could have one
container running a PostgreSQL database which serves data to another container
running a python environment to analyze the data:

@@

Unlike the much more heavyweight virtual machine approach, a single
computer can easily run 100s of such services each in their own
container. A rapidly expanding ecosystem of software around Docker,
such as fig ([github.com/docker/fig](https://github.com/docker/fig))
facilitates the complexity of running multiple containers.  This feature
making it easy to break computational elements down into logically
reusable chunks that come, batteries included, with everything they need
to run reproducibly.
